---
title: "Traumatic Brain Injury"
author: "Furkan Danisman"
date: "2024-03-23"
output:
  html_document:
    df_print: paged
---

# Functions and Definitions of the Part of the Brains

## Hippocampus-Left side

### Plays a crucial role in linguistic skills. In other words, episodic verbal memory.

## Capsules

### The internal capsule describes a region deep in the brain that functions as a communication pathway.

## Fornix Crux

### The fornix is a white matter tract that connects the hippocampus to several subcortical brain regions and is pivotal for episodic memory functioning.

## Gray Matter

### The place where the processing of sensation, perception, voluntary movement, learning, speech and cognition takes place.

```{r include=FALSE}
# Required Libraries
library(devtools)
library(car)
library(effects)
library(snow)
library(rbenchmark)
library(spida2)
library(readxl)
library(dplyr)
library(p3d)
library(lattice)
library(latticeExtra)
library(ggplot2)
library(gridExtra)
library(nlme)
library(rgl)
library(Hmisc)
```

### Wide-to-Long Form Data Transformation

```{r}
TBI <- read_excel("C:/Users/zenci/Downloads/TBI.xlsx")
TBI2 <- as.data.frame(TBI)

a <- names(TBI2)
names(TBI2) <- gsub("_(.)$","__\\1",a)
New_TBI <- tolong(TBI2,timevar = "visit",sep = "__",idvar="SubID")
New_TBI$SEX <- ifelse(New_TBI$sex==1,"Male","Female")
New_TBI_filtered <- New_TBI[complete.cases(New_TBI$SEX),]
```

### Time vs Hippocampus Left among each sex and group

```{r warning=FALSE}
HC_G_p <- ggplot(New_TBI_filtered, aes(x = date, y = HPC_L_TOT, group = SubID, col = SubID)) +
  geom_point() +
  geom_line() +
  facet_wrap(~group+SEX) +
  theme(legend.position = "none") +
  scale_y_continuous()

HC_G_p
```

### Hippocampus Left vs Fornix Cruxes

```{r warning=FALSE}
HC_F_L <- ggplot(New_TBI_filtered, aes(x = fornix_crux_L, y = HPC_L_TOT, group = SubID, col = SubID)) +
  geom_point() +
  geom_line() +
  facet_wrap(~SEX) +
  theme(legend.position = "none") +
  scale_y_continuous()

HC_F_R <- ggplot(New_TBI_filtered, aes(x = fornix_crux_R, y = HPC_L_TOT, group = SubID, col = SubID)) +
  geom_point() +
  geom_line() +
  facet_wrap(~SEX) +
  theme(legend.position = "none") +
  scale_y_continuous()

grid.arrange(HC_F_L,HC_F_R)

# Around the same relationship for both sides
```

### Hippocampus Left vs Capsule Anterior

```{r warning=FALSE , fig.height=12, fig.width=10}
HC_CA_L <- ggplot(New_TBI_filtered, aes(x = capsule_anterior_L, y = HPC_L_TOT, group = SubID, col = SubID)) +
  geom_point() +
  geom_line() +
  facet_wrap(~group+SEX) +
  theme(legend.position = "none") +
  scale_y_continuous()

HC_CA_R <- ggplot(New_TBI_filtered, aes(x = capsule_anterior_R, y = HPC_L_TOT, group = SubID, col = SubID)) +
  geom_point() +
  geom_line() +
  facet_wrap(~group+SEX) +
  theme(legend.position = "none") +
  scale_y_continuous()

grid.arrange(HC_CA_L,HC_CA_R)

```

### Hippocampus Left vs Age

```{r warning=FALSE}
HC_Age <- ggplot(New_TBI_filtered, aes(x = age, y = HPC_L_TOT, group = SubID, col = SubID)) +
  geom_point() +
  geom_line() +
  facet_wrap(~group+SEX) +
  theme(legend.position = "none") +
  scale_y_continuous()

HC_Age

# Similar to the time plot - just the scale is a bit different
```

### Hippocampus Left vs Gray Matter

```{r warning=FALSE}

HC_GM <- ggplot(New_TBI_filtered, aes(x = GM, y = HPC_L_TOT, group = SubID, col = SubID)) +
  geom_point() + 
  geom_line() +
  facet_wrap(~SEX) +
  theme(legend.position = "none") +
  scale_y_continuous()

HC_GM

# There is an observable positive relationship 
```

### Data Manipulations

```{r warning=FALSE}
# Data Manipulations 

New_TBI_filtered$SEX <- as.factor(New_TBI_filtered$SEX)
New_TBI_filtered$group <- as.factor(New_TBI_filtered$group)
clean_TBI_WO_Ls <- subset(New_TBI_filtered,!is.na(HPC_L_TOT) & !is.na(SEX) & !is.na(age_30))
clean_TBI_W_Ls <- subset(clean_TBI_WO_Ls, !is.na(fornix_crux_L) & !is.na(capsule_anterior_R) &
                         !is.na(GM))
only_clean_HPC <- subset(New_TBI_filtered,!is.na(HPC_L_TOT))

clean_TBI_W_Ls$visit<- as.factor(clean_TBI_W_Ls$visit)
clean_TBI_W_Ls$SubID <- as.factor(clean_TBI_W_Ls$SubID)
clean_TBI_W_Ls$id <- as.factor(clean_TBI_W_Ls$id)
a <- select(clean_TBI_W_Ls,c(HPC_L_TOT,HPC_R_TOT,age_30,fornix_crux_L,capsule_anterior_R,GM,visit,SEX,group))
b <- subset(clean_TBI_W_Ls,visit==2)
c <- subset(b,age>25 & age< 50)
d <- subset(b,age<=25)
middle_age_prop <- nrow(c)/nrow(b)
young_age_prop <- nrow(d)/nrow(b)
old_age_prop <- (nrow(b) - nrow(c) - nrow(d)) / nrow(b)
Prop_age <- data.frame("Younger -25" = young_age_prop, "Middle Age 25-50"=middle_age_prop,
                       "Older 50-"=old_age_prop)
rownames(Prop_age) <- "Proportion"
colnames(Prop_age) <- c("Younger-25","Middle Age 25-50", "Older 50-")
Prop_age <- Prop_age *100
range_of_age <- range(clean_TBI_W_Ls$age_30)
```

### Age Proportion

```{r warning=FALSE}
Prop_age
```

### QQ-plot For each X Variables

```{r}
xqplot(a)

```

### Models

```{r}
# Mixed Models 

# Model with the random part

fit.wLs <-lme( HPC_L_TOT ~ SEX + age_30 +group*age_30 + capsule_anterior_R +
                 GM  + fornix_crux_L, clean_TBI_W_Ls, 
               random = ~ 1 + age_30 | SubID, correlation = corCAR1(form = ~ age_30 | SubID) )

# Model without the random part

fit.wLs2 <- gls( HPC_L_TOT ~ SEX + age_30 + group*age_30 + capsule_anterior_R +GM  + fornix_crux_L, clean_TBI_W_Ls )

# Testing G-matrix 

anova(fit.wLs,fit.wLs2)

# G-matrix is significantly greater than 0 

# Model Updates to compare models - Section 1 

fit2 <- update( fit.wLs, . ~ . + SEX*age_30)
fit4 <- update( fit2, . ~ . + cvar(GM,SubID)) # To factor in variability between subjects
fit2l.1 <- fit4
```

### Diagnostic for heteroskedasticity ----

```{r}
plot( fit2l.1, sqrt( abs( resid(.))) ~ fitted(.), id = .02) # exploratory version

# fancy version
plot( fit2l.1, sqrt( abs( resid(.))) ~ fitted(.), id = .02)

plot( fit2l.1, sqrt( abs( resid(., type = 'p'))) ~ fitted(.) | SEX, id = .03,
      panel = function(x, y, ...) {
        panel.xyplot( x, y, ...)
        panel.loess( x, y,...)
      })

# id didn't work so we do it manually

#trellis.focus()            
#panel.identify(labels = clean_TBI_W_Ls$SubID)
#trellis.unfocus()
```

### Identified Points: SubID = { 368 - 380 - 325 - 462 - 346 - 453 - 358 }

```{r}
sus <- c(368,380,325,462,346,453,358)
subset_data <- subset(clean_TBI_W_Ls,SubID %in% sus)

```

### Plotting the highlighted "suspicious patients".

```{r}
HC_Sample2.1 <- ggplot(New_TBI_filtered, aes(x = date, y = HPC_L_TOT, group = SubID, col = SubID)) +
  geom_point() +
  geom_line() +
  facet_wrap(~SEX) +
  theme(legend.position = "none") +
  geom_text(data = subset(subset_data, SubID %in% sus),
            aes(label = SubID), vjust = -0.5) + 
  scale_y_continuous()

HC_sample2 <- ggplot(data = subset_data,aes(x=date,y=HPC_L_TOT,col=SubID,group=SubID))+
  geom_point()+
  geom_line()+
  theme(legend.position="none")+
  geom_text(data = subset(subset_data, SubID %in% sus),
            aes(label = SubID), vjust = -0.5)+
  scale_y_continuous()

grid.arrange(HC_Sample2.1,HC_sample2)

possible_rmv <- c(346)
```

### QQnorms

```{r}

qqnorm( fit2l.1, id = .05 )
qqnorm( fit2l.1, ~ resid(.) | SEX, id = .05 )

```

### Acting on diagnostics

### Should we drop some observations?

```{r}
possible_rmv <- c(346)

fit2l.d <- update( fit2l.1, data = subset(clean_TBI_W_Ls, !(id %in% possible_rmv)) )

# or

fit2l.d <- update( fit2l.1, subset = !(id %in% possible_rmv) )
```

### Heteroscedasticity revisited

```{r}

plot( fit2l.d, sqrt( abs( resid(., type = 'p'))) ~ fitted(.), id = .03)

plot( fit2l.d, sqrt( abs( resid(., type = 'p'))) ~ fitted(.), id = .03,
      panel = function(x, y, ...) {
        panel.xyplot( x, y, ...)
        panel.loess( x, y,...)
      })

plot( fit2l.d, sqrt( abs( resid(., type = 'p'))) ~ fitted(.) | SEX, id = .03,
      panel = function(x, y, ...) {
        panel.xyplot( x, y, ...)
        panel.loess( x, y,...)
      })

```

### QQnorms

```{r}

qqnorm( fit2l.d, id = .05 )
qqnorm( fit2l.d, ~ resid(.) | SEX, id = .05 )

```

### Variogram: diagnostics for autocorrelation

```{r}
vv <- Variogram( fit2l.d , form = ~ age_30 | SubID, maxdist = .7)
plot(vv)

# Differences between pairs of close residuals have smaller variance than between far apart ones.

```

### Model Comparison 

```{r}
w.fit <- wald(fit.wLs)[[1]]
a <- w.fit$anova
w.fit2 <- wald(fit2)[[1]]
b <- w.fit2$anova
w.fit4 <- wald(fit4)[[1]]
c <- w.fit4$anova
w.fitd <- wald(fit2l.d)[[1]]
d <- w.fitd$anova
a <- as.matrix(a)
b <- as.matrix(b)
c <- as.matrix(c)
d <- as.matrix(d)
e <- cbind(a,b,c,d)
colnames(e) <- c("Fit","Fit-Interaction","Fit-Contextual","Fit-W/outlier")
e

# The p-value for all of them is relatively low. We can continue with the final model.  

```

## Final Model

### Y variable: Left-Hippocampus-Total (Volume)

### X variables:

### Fixed Part: Sex + Age (Centered at 30) + Sex\*Age (Centered at 30) + Group\*Age (Centered at 30) + Group + Capsule Anterior-R + Gray Matter + Fornix Crux - L + cvar(Gray Matter)

### Random Part: Intercept + Age_30 conditioned on subject

### Correlation: \~ Age_30 \| Subject

## Questions

### Q1 - Does the hippocampus volume undergo significant changes after an injury, and do these changes vary over time post-injury?

```{r warning=FALSE}
# Answer1: Since we do not have pre-injury data, we can not conclude if the changes are significant or natural.  

# Answer2: However, we can observe the pattern after the injury, factoring in other variables. For example, we can observe how different the patients are compared to the control group. There appears to be a change in the slope for the two groups. 

# Plot vs time wrapped by group and sex
HC_G_p # Let's revisit the graph

# Wald-Test

wald(fit2l.d,"group")

# Also, there seems to be a significant difference of 510 Hippocampus (TOT) in the intercept between patients and the control group. However, for the slope, there is a small difference of 1.89 for the patients. Although difference in slope itself is not significant, at d=2 level group variable is highly significant with the p-value of 0.007 

```

### Q2 - At different time points post-injury, does the level of cognitive recovery (especially language comprehension ability) improve over age or time?

```{r warning=FALSE}
HC_Age

# Based on the age graph, there could be a slight downward pattern for the intercept as age increases for Males. 

wald(fit2l.d,"age_30")


# The relationship of age differs by sex with the interaction term. 
# Although age and interaction of age with group is not significant, we can not drop them as d=3 level they are significant together. Slope of age highly differs between male and female. 

# Auto-correlation 

intervals(fit2l.d)

# As phi is significantly positive (do not include 0), we can conclude that there is a positive autocorrelation. 
# In other words, if the hippocampus level goes down, it will likely hold that trajectory for the following days, and vice versa. We know that the trend of time might differ among sexes. 

# For female patients, the hippocampus total increases by 3.94 for each increase in age, holding others constant. 

# For males, the hippocampus total decreases by 11.52 for each increase in age, holding others constant. 

# Therefore, combined with the positive autocorrelation, the trend appears to be upward for females as a positive trajectory follows a positive trajectory. Meanwhile, for males, the trend is downward, as the negative track follows the negative track. 

```
### G-Matrix

```{r}
getVarCov(fit2l.d)

# The covariance is negative between intercept and slope which indicates a negative relationship between the two. 
```

### Intercept vs Age among each sex

```{r}
data = subset(clean_TBI_W_Ls, !(id %in% possible_rmv))
RE <- cbind( ranef( fit2l.d),up(data,~SubID) )
colnames(RE)[2] <- "Slope"
RE = subset(RE, !(id %in% 453))
ggplot(RE, aes(x = Slope, y = `(Intercept)`,group = SEX,col=SEX)) +
  facet_wrap(~SEX)+
  geom_point() +  # Plot the data points
  geom_smooth(method = "lm", se = FALSE)

# As age increases, there seems to be a pattern for upward intercept for males, whereas, for females, it is relatively straight.

```

### QQ-Norm

```{r}
qqnorm(RE[[1]])
qqnorm(RE[[2]])
```


### Q3 - During the process of changes in hippocampal volume post-injury, does sex impact this process?

```{r warning=FALSE}

wald(fit2l.d,"SEX")

# Suggests sex and interaction together are significant at 84% confidence level.

# The intercept is significantly higher for males with the value of 170. However, this relationship opposite for the slopes. 


```

### Q4 - Is the deep communication pathway region and episodic memory of the brain, capsule anterior and fornix crux, respectively, related to decreased hippocampal volume and reduced cognitive function?

```{r warning=FALSE}
HC_CA_R
HC_F_L


wald(fit2l.d,"fornix_crux_L")
wald(fit2l.d,"capsule_anterior_R")

#(MORE TO TELL IN THE REPORT - IT IS JUST A REPORT ON THE VALUE) 

```

### Q5 - Considering that the right side of the hippocampus is related to spatial perception abilities, is there a connection between spatial perception abilities and language comprehension and production?

```{r warning=FALSE}
HC_R <- ggplot(New_TBI_filtered, aes(x = HPC_R_TOT, y = HPC_L_TOT, group = SubID, col = SubID)) +
  geom_point() + 
  geom_line() +
  facet_wrap(~group+SEX) +
  theme(legend.position = "none") +
  scale_y_continuous()

HC_R 

# As expected, there is a strong relationship, but we suspect it is a confounding variable. But we can not measure it. 
# Hence, it is better to avoid adding as a variable at this point. 


```

### Q6 - What is the connection between the Gray Matter, where processing sensation, perception, voluntary movement, learning, speech and cognition happens, and the hippocampus left-side, where the brain constructs the languages?

```{r warning=FALSE}

wald(fit2l.d,"GM")

# (MORE TO TELL IN THE REPORT - IT IS JUST A REPORT ON THE VALUE) It appears to play a significant role in predicting hippocampus level. It is highly significant with a low estimate value compared to other units.  

```

### Q7 - What would be the predicted Hippocampus level for a male at the age of 20 with a traumatic brain injury who has an average fornix crux, capsule anterior, and Gray Matter Level?

```{r warning=FALSE}
a <- select(clean_TBI_W_Ls,id,fornix_crux_L,capsule_anterior_R,GM)
b <- a %>% group_by(id)
b <- a %>% group_by(id) %>% slice(1)
L <- c(1,1,-10,1,mean(clean_TBI_W_Ls$capsule_anterior_R),mean(clean_TBI_W_Ls$GM),mean(clean_TBI_W_Ls$fornix_crux_L),
  c(mean(b$fornix_crux_L),mean(b$capsule_anterior_R)),-10)
L <- as.matrix(L)
L <- t(L)
# wald(fit2l.d,L)
```

### How about a female with the same features?

```{r}
L2 <- c(1,0,-10,1,mean(clean_TBI_W_Ls$capsule_anterior_R),mean(clean_TBI_W_Ls$GM),mean(clean_TBI_W_Ls$fornix_crux_L),c(mean(b$fornix_crux_L),mean(b$capsule_anterior_R)),0)
L2 <- as.matrix(L2)
L2 <- t(L2)
L3 <- rbind(L,L2)
rownames(L3) <- c("Male","Female")
Young_result <- wald(fit2l.d,L3)

```

### How about middle-aged (50) people?

```{r}
L <- c(1,1,20,1,mean(clean_TBI_W_Ls$capsule_anterior_R),mean(clean_TBI_W_Ls$GM),mean(clean_TBI_W_Ls$fornix_crux_L),
       c(mean(b$fornix_crux_L),mean(b$capsule_anterior_R)),20)
L <- as.matrix(L)
L <- t(L)

#wald(fit2l.d,L)

# Female 

L2 <- c(1,0,20,1,mean(clean_TBI_W_Ls$capsule_anterior_R),mean(clean_TBI_W_Ls$GM),mean(clean_TBI_W_Ls$fornix_crux_L),
        c(mean(b$fornix_crux_L),mean(b$capsule_anterior_R)),0)
L2 <- as.matrix(L2)
L2 <- t(L2)
L3 <- rbind(L,L2)
rownames(L3) <- c("Male","Female")
Middle_aged_result <- wald(fit2l.d,L3)
```

### How about someone in their 70s?

```{r}

L <- c(1,1,44,1,mean(clean_TBI_W_Ls$capsule_anterior_R),mean(clean_TBI_W_Ls$GM),mean(clean_TBI_W_Ls$fornix_crux_L),
       c(mean(b$fornix_crux_L),mean(b$capsule_anterior_R)),44)
L <- as.matrix(L)
L <- t(L)


# How about a female with same features?

L2 <- c(1,0,40,1,mean(clean_TBI_W_Ls$capsule_anterior_R),mean(clean_TBI_W_Ls$GM),mean(clean_TBI_W_Ls$fornix_crux_L),
        c(mean(b$fornix_crux_L),mean(b$capsule_anterior_R)),0)
L2 <- as.matrix(L2)
L2 <- t(L2)
L3 <- rbind(L,L2)
rownames(L3) <- c("Male","Female")
Old_result <-wald(fit2l.d,L3)

# Age Comparison 

a <- list()
a[[1]] <- Young_result
a[[2]] <- Middle_aged_result
a[[3]] <- Old_result
names(a) <- c("Young-20's","Middle-50's","Elder-70's")
a

# As males get older, we can observe an increase in the estimate, from around 4500's to 4000's and 3800's. However, for females, the rate of change does not look as dramatic, in the contrary, it even increases. 

# A female subject on her 70's have a better hippocampus level than a male on his 50's after the TBI.
```

### Q8: Visit and improvement - Do people who show improvement still visit?

```{r}
table(only_clean_HPC$visit)
v4 <- subset(only_clean_HPC,visit==4)
v3 <- subset(only_clean_HPC,visit==3)
v2 <- subset(only_clean_HPC,visit==2)

v41 <- select(v4,id,HPC_L_TOT)
v31 <- select(v3,id,HPC_L_TOT)
v21 <- select(v2,id,HPC_L_TOT)
id_list <- 0

diffHC <- 0
for (i in 1:nrow(v3)) {
  if(v31[i,"id"]%in%v21$id){
    id_list[i] <- as.character(v31[i,"id"]) 
    diffHC[i] <- v21[which(v21$id==v31[i,"id"]),"HPC_L_TOT"] - v31[i,"HPC_L_TOT"]
  }else{id_list[i] <- 0}
}

id_list2 <- 0
for (i in 1:nrow(v4)) {
  if(!(v41[i,"id"]%in%v31$id)){
    id_list2[i] <- as.character(v31[i,"id"]) 
  }else{id_list2[i] <- 0}
}

id_list[which(diffHC<0)] %in% id_list2[id_list2!=0]
id_list[which(diffHC<0)] 

# All of the subjects in the patient's group, 6 out of 6, who had an improvement, stopped coming after the visit-3. In contrast, only a tiny percentage of the non-improved patients stopped coming.  

```


